The Efficient Handling of Guards \\in the Design of RPython's Tracing JIT

### 2012-08-16 talk/vmil2012/paper.tex
include Sven's notes

\title{The Efficient Handling of Guards in the Design of RPython's Tracing JIT}

@@

%Section~\ref{sec:Evaluation} presents Figures about the absolute number of
%operations for each benchmark, and the overhead produced by the information
%stored at the different levels for the guards
In this paper we want to substantiate the aforementioned observations about guards and
describe based on them the reasoning behind their implementation in
RPython's tracing just-in-time compiler. the contributions of this paper are:
\begin{itemize}
  \item an analysis and benchmark of guards in the context of RPython's tracing JIT,
  %An analysis of guards in the context of RPython's tracing JIT to
  %substantiate the aforementioned observation, based on a set of benchmarks,
  \item detailed measurements about the frequency and the
  overhead associated with guards, and
  \item a description about how guards are implemented in the high\-
  and low-level components of the JIT and describe the rationale behind the design
\end{itemize}

The set of central concepts upon which this work is based are described in
Section~\ref{sec:Background}, such as the PyPy project, the RPython language
and its meta-tracing JIT. Based on these concepts in Section~\ref{sec:Resume
Data} we proceed to describe for RPython's tracing JIT the details of guards in
the frontend. In this context the frontend is concerned with recording and storing the
information required to rebuild the interpreter state in case of a guard
failure. Once the frontend has traced and optimized a loop it invokes the
backend to compile the operations to machine code, Section~\ref{sec:Guards in
the Backend} describes the low-level aspects of how guards are implemented in
the machine specific JIT-backend. The frequency of guards and the overhead associated with the
implementation described in this paper is discussed in
Section~\ref{sec:evaluation}. Section~\ref{sec:Related Work} presents an
overview about how guards are treated in the context of other just-in-time
compilers. Finally Section~\ref{sec:Conclusion} summarizes our conclusions and
gives an outlook on further research topics.

@@

RPython, the language and the toolset originally created to implement the
Python interpreter have developed into a general environment for experimenting
and developing fast and maintainable dynamic language implementations. Besides
the Python interpreter there are several experimental language implementation at different
levels of completeness, e.g. for Prolog~\cite{bolz_towards_2010}, Smalltalk~\cite{bolz_towards_2010}, JavaScript and R.

@@

\begin{figure}
    \input{figures/example.tex}
    \caption{Example Program}
    \label{fig:example}
\end{figure}

@@

The core idea of storing resume data as compactly as possible
is to share parts of the data structure between subsequent guards.
This is useful because the density of guards in traces is so high,
that quite often not much changes between them.
Since resume data is a linked list of symbolic frames
in many cases only the information in the top frame changes from one guard to the next.
The other symbolic frames can often just be reused.
The reason for this is that during tracing only the variables
of the currently executing frame can change.
Therefore if two guards are generated from code in the same function
the resume data of the rest of the frame stack can be reused.

In addition to sharing as much as possible between subsequent guards
a compact representation of the local variables of symbolic frames is used.
Every variable in the symbolic frame is encoded using two bytes.
Two bits are used as a tag to denote where the value of the variable
comes from.
The remaining 14 bits are a payload that depends on the tag bits.

@@

After the recorded trace has been optimized it is handed over to the platform specific
backend to be compiled to machine code. The compilation phase consists of two
passes over the lists of instructions, a backwards pass to calculate live
ranges of IR-level variables and a forward pass to emit the instructions. During
the forward pass IR-level variables are assigned to registers and stack
locations by the register allocator according to the requirements of the
emitted instructions.  Eviction/spilling is performed based on the live range
information collected in the first pass. Each IR instruction is transformed
into one or more machine level instructions that implement the required
semantics. Operations without side effects whose result is not used are not
emitted. Guards instructions are transformed into fast checks at the machine
code level that verify the corresponding condition.  In cases the value being
checked by the guard is not used anywhere else the guard and the operation
producing the value can merged, further reducing the overhead of the guard.
Figure~\ref{fig:trace-compiled} shows how the \lstinline{int_eq} operation
followed by a \lstinline{guard_false} from the trace in Figure~\ref{fig:trace-log} are compiled to
pseudo-assembler if the operation and the guard are compiled separated or if
they are merged.

@@

First a special data
structure called \emph{backend map} is created. This data structure encodes the
mapping from IR-variables needed by the guard to rebuild the state to the
low-level locations (registers and stack) where the corresponding values will
be stored when the guard is executed.
This data
structure stores the values in a succinct manner using an encoding that requires
8 bits to store 7 bits of information, ignoring leading zeros. This encoding is efficient to create and
provides a compact representation of the needed information in order
to maintain an acceptable memory profile.

Second for each guard a piece of code is generated that acts as a trampoline.
Guards are implemented as a conditional jump to this trampoline in case the
guard check fails.
In the trampoline the pointer to the
backend map is loaded and after storing the current execution state
(registers and stack) execution jumps to a generic bailout handler, also known
as \emph{compensation code},
that is used to leave the compiled trace.

@@

As in previous sections the underlying idea for the low-level design of guards is to have
a fast on-trace profile and a potentially slow one in case
the execution has to return to the interpreter. At the same
time the data stored in the backend, required to rebuild the state, should be as
compact as possible to reduce the memory overhead produced by the large number
of guards, the numbers in Figure~\ref{fig:backend_data} illustrate that the
compressed encoding currently has about 15\% to 25\% of the size of of the
generated instructions on x86.

@@

Tracing JIT compilers only compile the subset of the code executed in a program
that occurs in a hot loop, for this reason the amount of generated machine
code will be smaller than in other juts-in-time compilation approaches.  This
creates a larger discrepancy between the size of the resume data when
compared to the size of the generated machine code and illustrates why it is important to compress the resume data information.

@@

\section*{Acknowledgements}
We would like to thank David Edelsohn, Samuele Pedroni and Stephan Zalewski for their helpful
feedback and valuable comments while writing this paper.
We thank the PyPy and RPython community for their continuous support and work:
Armin Rigo, Antonio Cuni, Maciej Fijałkowski, Samuele Pedroni, and countless
others. Any remaining errors are our own.

%\section*{Appendix}
%\todo{remove this section and the figures}
%\begin{figure*}
%    \include{figures/ops_count_table}
%    \caption{Relative numbers of operations in the traces generated for
%    different benchmarks}
%    \label{fig:ops_count}
%\end{figure*}
%\begin{figure*}
%\centering
%\includegraphics[width=\textwidth]{figures/ops_pie.pdf}
%\caption{Relative frequency of operations before and after optimization}
%\label{fig:ops_pie}
%\end{figure*}
\bibliographystyle{abbrv}
\bibliography{zotero,paper}
\listoftodos
\end{document}

>>>>

\title{The Efficient Handling of Guards \\in the Design of RPython's Tracing JIT}

@@

%Section~\ref{sec:Evaluation} presents Figures about the absolute number of
%operations for each benchmark, and the overhead produced by the information
%stored at the different levels for the guards
In this paper we want to substantiate the aforementioned observations about guards and
describe based on them the reasoning behind their implementation in
RPython's tracing just-in-time compiler. The contributions of this paper are:
\begin{itemize}
  \item an analysis and benchmark of guards in the context of RPython's tracing JIT,
  %An analysis of guards in the context of RPython's tracing JIT to
  %substantiate the aforementioned observation, based on a set of benchmarks,
  \item detailed measurements about the frequency and the
  overhead associated with guards, and
  \item a description about how guards are implemented in the high\-
  and low-level components of the JIT and describe the rationale behind the design
\end{itemize}

The set of central concepts upon which this work is based are described in
Section~\ref{sec:Background}, such as the PyPy project, the RPython language
and its meta-tracing JIT. Based on these concepts in Section~\ref{sec:Resume
Data} we proceed to describe for RPython's tracing JIT the details of guards in
the frontend. In this context the frontend is concerned with recording and storing the
information required to rebuild the interpreter state in case of a guard
failure. Once the frontend has traced and optimized a loop it invokes the
backend to compile the operations to machine code, Section~\ref{sec:Guards in
the Backend} describes the low-level aspects of how guards are implemented in
the machine specific JIT-backend. The frequency of guards and the overhead associated with the
implementation described in this paper is discussed in
Section~\ref{sec:evaluation}. Section~\ref{sec:Related Work} presents an
overview about how guards are treated in the context of other just-in-time
compilers. Finally, Section~\ref{sec:Conclusion} summarizes our conclusions and
gives an outlook on further research topics.

@@

RPython, the language and the toolset originally created to implement the
Python interpreter have developed into a general environment for experimenting
and developing fast and maintainable dynamic language implementations. Besides
the Python interpreter there are several experimental language implementation at different
levels of completeness, e.g. for Prolog~\cite{bolz_towards_2010}, Smalltalk~\cite{bolz_back_2008}, JavaScript and R.

@@

\begin{figure}
    \input{figures/example.tex}
    \caption{Example program}
    \label{fig:example}
\end{figure}

@@

The core idea of storing resume data as compactly as possible
is to share parts of the data structure between subsequent guards.
This is useful because the density of guards in traces is so high,
that quite often not much changes between them.
Since resume data is a linked list of symbolic frames,
in many cases only the information in the top frame changes from one guard to the next.
The other symbolic frames can often be reused.
The reason for this is that, during tracing only the variables
of the currently executing frame can change.
Therefore if two guards are generated from code in the same function
the resume data of the rest of the frame stack can be reused.

In addition to sharing as much as possible between subsequent guards,
a compact representation of the local variables of symbolic frames is used.
Every variable in the symbolic frame is encoded using two bytes.
Two bits are used as a tag to denote where the value of the variable
comes from.
The remaining 14 bits are a payload that depends on the tag bits.

@@

After the recorded trace has been optimized, it is handed over to the platform specific
backend to be compiled to machine code. The compilation phase consists of two
passes over the lists of instructions, a backwards pass to calculate live
ranges of IR-level variables and a forward pass to emit the instructions. During
the forward pass IR-level variables are assigned to registers and stack
locations by the register allocator according to the requirements of the
emitted instructions.  Eviction/spilling is performed based on the live range
information collected in the first pass. Each IR instruction is transformed
into one or more machine level instructions that implement the required
semantics. Operations without side effects whose result is not used are not
emitted. Guard instructions are transformed into fast checks at the machine
code level that verify the corresponding condition.  In cases the value being
checked by the guard is not used anywhere else the guard and the operation
producing the value can merged, further reducing the overhead of the guard.
Figure~\ref{fig:trace-compiled} shows how the \lstinline{int_eq} operation
followed by a \lstinline{guard_false} from the trace in Figure~\ref{fig:trace-log} are compiled to
pseudo-assembler if the operation and the guard are compiled separated or if
they are merged.

@@

First, a special data
structure called \emph{backend map} is created. This data structure encodes the
mapping from IR-variables needed by the guard to rebuild the state to the
low-level locations (registers and stack) where the corresponding values will
be stored when the guard is executed.
This data
structure stores the values in a succinct manner using an encoding that requires
8 bits to store 7 bits of information, ignoring leading zeros. This encoding is efficient to create and
provides a compact representation of the needed information in order
to maintain an acceptable memory profile.

Second, for each guard a piece of code is generated that acts as a trampoline.
Guards are implemented as a conditional jump to this trampoline in case the
guard check fails.
In the trampoline, the pointer to the
backend map is loaded and after storing the current execution state
(registers and stack) execution jumps to a generic bailout handler, also known
as \emph{compensation code},
that is used to leave the compiled trace.

@@

As in previous sections, the underlying idea for the low-level design of guards is to have
a fast on-trace profile and a potentially slow one in case
the execution has to return to the interpreter. At the same
time, the data stored in the backend, required to rebuild the state, should be as
compact as possible to reduce the memory overhead produced by the large number
of guards. The numbers in Figure~\ref{fig:backend_data} illustrate that the
compressed encoding currently has about 15\% to 25\% of the size of of the
generated instructions on x86.

@@

Tracing JIT compilers only compile the subset of the code executed in a program
that occurs in a hot loop, for this reason the amount of generated machine
code will be smaller than in other just-in-time compilation approaches.  This
creates a larger discrepancy between the size of the resume data when
compared to the size of the generated machine code and illustrates why it is important to compress the resume data information.

@@

\section*{Acknowledgements}
We would like to thank David Edelsohn, Samuele Pedroni, Stephan Zalewski and Sven Hager for their helpful
feedback and valuable comments while writing this paper.
We thank the PyPy and RPython community for their continuous support and work:
Armin Rigo, Antonio Cuni, Maciej Fijałkowski, Samuele Pedroni, and countless
others. Any remaining errors are our own.

%\section*{Appendix}
%\todo{remove this section and the figures}
%\begin{figure*}
%    \include{figures/ops_count_table}
%    \caption{Relative numbers of operations in the traces generated for
%    different benchmarks}
%    \label{fig:ops_count}
%\end{figure*}
%\begin{figure*}
%\centering
%\includegraphics[width=\textwidth]{figures/ops_pie.pdf}
%\caption{Relative frequency of operations before and after optimization}
%\label{fig:ops_pie}
%\end{figure*}
\bibliographystyle{abbrv}
\bibliography{zotero,paper}
%\listoftodos
\end{document}
